---
title: "DVP_required_viz"
author: "Ankit"
date: "2024-11-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Make relative paths work when rendering from this report/ folder
knitr::opts_knit$set(root.dir = "..")
```

## Loading necessary libraries and checking the structure of the tables.

```{r}
library(dplyr)
library(ggplot2)
balls<- read.csv("data/deliveries.csv")
matches<-read.csv("data/matches.csv")
# Add a new column 'over' by extracting the integer part of 'ball'
balls$over <- floor(balls$ball)
str(balls)
str(matches)

```
## Aggregating the values for runs/scored per over for a team

```{r}
# Aggregate average runs scored per team and per over
team_runs_per_over <- balls %>%
  group_by(batting_team, over) %>%
  summarise(average_runs = mean(runs_off_bat, na.rm = TRUE)) %>%
  ungroup()

# Check the result
head(team_runs_per_over)
```
## Converting values of runs scored per over per team into wide format to view over wise run distribution throughout the tournament.

```{r}
# Reshape the data: wide format where each column is the runs for each over
team_runs_wide <- team_runs_per_over %>%
  tidyr::pivot_wider(names_from = over, values_from = average_runs, values_fill = 0)
```


## Aggregating the values for wickets/taken per over for a team
```{r}
# Step 1: Count wickets per over for each bowling team
team_wickets_per_over <- balls %>%
  filter(!is.na(wicket_type) & wicket_type != "") %>%
  group_by(bowling_team, over) %>%
  summarise(wickets = n(), .groups = 'drop')

# Step 2: Calculate the number of matches each bowling team played
matches_per_team <- balls %>%
  group_by(bowling_team, match_id) %>%
  summarise() %>%
  group_by(bowling_team) %>%
  summarise(matches_played = n())

# Step 3: Join matches_played with team_wickets_per_over and calculate average wickets per over
team_wickets_avg_per_over <- team_wickets_per_over %>%
  left_join(matches_per_team, by = "bowling_team") %>%
  mutate(average_wickets = wickets / matches_played) %>%
  select(bowling_team, over, average_wickets)

# Reshape to wide format
team_wickets_wide <- team_wickets_avg_per_over %>%
  tidyr::pivot_wider(names_from = over, values_from = average_wickets, values_fill = 0)

# Display the result
print(team_wickets_wide)

```


## Calculate runs_per_match

```{r}
# Group by match_id and batting_team to get total runs per match
runs_per_match <- balls %>%
  group_by(match_id, batting_team) %>%
  summarise(total_runs = sum(runs_off_bat + extras, na.rm = TRUE)) %>%
  ungroup()

# Calculate average runs per match for each team
avg_runs_per_team <- runs_per_match %>%
  group_by(batting_team) %>%
  summarise(avg_runs = mean(total_runs, na.rm = TRUE)) %>%
  rename(team = batting_team)

```
## Calculate dot_balls_faced

```{r}
# Filter for dot balls (runs_off_bat == 0 and extras == 0)
dot_balls_faced <- balls %>%
  filter(runs_off_bat == 0 & extras == 0) %>%
  group_by(match_id, batting_team) %>%
  summarise(dot_balls = n()) %>%
  ungroup()

# Calculate average dot balls faced per match for each team
avg_dot_balls_per_team <- dot_balls_faced %>%
  group_by(batting_team) %>%
  summarise(avg_dot_balls = mean(dot_balls, na.rm = TRUE)) %>%
  rename(team = batting_team)

```
## Calculate wickets_per_match

```{r}
# Filter rows where there was a wicket (wicket_type not NA or empty)
wickets_per_match <- balls %>%
  filter(!is.na(wicket_type) & wicket_type != "") %>%
  group_by(match_id, bowling_team) %>%
  summarise(total_wickets = n()) %>%
  ungroup()

# Calculate average wickets taken per match for each team
avg_wickets_per_team <- wickets_per_match %>%
  group_by(bowling_team) %>%
  summarise(avg_wickets = mean(total_wickets, na.rm = TRUE)) %>%
  rename(team = bowling_team)

```


## Calculate extras_per_match
```{r}
# Group by match_id and bowling_team to get total extras conceded per match
extras_per_match <- balls %>%
  group_by(match_id, bowling_team) %>%
  summarise(total_extras = sum(extras, na.rm = TRUE)) %>%
  ungroup()

# Calculate average extras conceded per match for each team
avg_extras_per_team <- extras_per_match %>%
  group_by(bowling_team) %>%
  summarise(avg_extras = mean(total_extras, na.rm = TRUE)) %>%
  rename(team = bowling_team)

```


## Calculating wins/team in the tournament
```{r}
team_wins <- matches %>%
  filter(winner != "") %>%   # Exclude matches without a winner (e.g., ties or abandoned)
  group_by(winner) %>%
  summarise(wins = n()) %>%
  rename(team = winner)

```



##  Combine all the tables to create team_performance

```{r}
# Assuming `team_wins` is the table with columns 'team' and 'wins'
team_performance <- avg_runs_per_team %>%
  left_join(avg_dot_balls_per_team, by = "team") %>%
  left_join(avg_wickets_per_team, by = "team") %>%
  left_join(avg_extras_per_team, by = "team") %>%
  left_join(team_wins, by = "team") # 'team_wins' contains team and wins columns

```


```{r}
# Filter for boundaries (4s and 6s) and count them per match
boundaries_per_match <- balls %>%
  filter(runs_off_bat == 4 | runs_off_bat == 6) %>%
  group_by(match_id, batting_team) %>%
  summarise(boundaries = n()) %>%
  ungroup()

# Calculate average boundaries per match for each team
avg_boundaries_per_team <- boundaries_per_match %>%
  group_by(batting_team) %>%
  summarise(avg_boundaries = mean(boundaries, na.rm = TRUE)) %>%
  rename(team = batting_team)

```
```{r}
team_performance <- team_performance %>%
  left_join(avg_boundaries_per_team, by = "team")

```


```{r}
str(team_performance)
head(team_performance)
```
## Calculating ties and losses per team

```{r}
# Base set of teams
all_teams <- team_performance %>% distinct(team)

# Losses: count when a team appears in the match but is NOT the winner.
# (This ignores abandoned/no-result matches where winner is blank.)
losses <- matches %>%
  filter(winner != "") %>%
  mutate(
    team1_loss = (team1 != winner),
    team2_loss = (team2 != winner)
  ) %>%
  summarise(
    losses_team1 = sum(team1_loss, na.rm = TRUE),
    losses_team2 = sum(team2_loss, na.rm = TRUE)
  )

# Compute losses per team from match rows
losses <- matches %>%
  filter(winner != "") %>%
  transmute(team = team1, loss = as.integer(team1 != winner)) %>%
  bind_rows(matches %>% filter(winner != "") %>% transmute(team = team2, loss = as.integer(team2 != winner))) %>%
  group_by(team) %>%
  summarise(losses = sum(loss, na.rm = TRUE), .groups = 'drop')

# Ties / no-result (if present in the dataset): treat rows with blank winner as tie/no-result
# Note: Some datasets encode this differently; this provides a safe default.
ties <- matches %>%
  filter(winner == "") %>%
  transmute(team = team1, tie = 1L) %>%
  bind_rows(matches %>% filter(winner == "") %>% transmute(team = team2, tie = 1L)) %>%
  group_by(team) %>%
  summarise(ties = sum(tie, na.rm = TRUE), .groups = 'drop')

# Ensure all teams exist with 0 defaults
losses <- all_teams %>% left_join(losses, by = 'team') %>% mutate(losses = ifelse(is.na(losses), 0L, losses))
ties   <- all_teams %>% left_join(ties,   by = 'team') %>% mutate(ties   = ifelse(is.na(ties),   0L, ties))

# Join ties/losses into team_performance
team_performance <- team_performance %>%
  left_join(losses, by = 'team') %>%
  left_join(ties, by = 'team')
```

```{r}
str(team_performance)
```

## Calculating PCA for KPI's

```{r}
# Load necessary libraries
library(dplyr)
library(ggplot2)

# Select KPIs and scale them
kpi_data <- team_performance %>%
  select(avg_runs, avg_wickets, avg_dot_balls, avg_extras, avg_boundaries) %>%
  scale()

# Perform PCA
pca_result <- prcomp(kpi_data, center = TRUE, scale. = TRUE)

# Add PCA scores and wins to a new data frame
pca_scores <- as.data.frame(pca_result$x) %>%
  mutate(team = team_performance$team,
         wins = team_performance$wins)
```

## PC1 vs Wins

```{r}
# Scatter plot of PC1 vs Wins
ggplot(pca_scores, aes(x = wins, y = PC1)) +
  geom_point(size = 4, color = "blue") +
  geom_text(aes(label = team), vjust = -1, size = 3) +
  labs(title = "PC1 of Team Performance KPIs vs Wins",
       x = "Wins",
       y = "Principal Component 1 (PC1)") +
  theme_minimal()

```


## PC1 vs PC2 colur coded by wins
```{r}
# Scatter plot of PC1 vs PC2, colored by wins
ggplot(pca_scores, aes(x = PC1, y = PC2, color = wins)) +
  geom_point(size = 4) +
  geom_text(aes(label = team), vjust = -1, size = 3) +
  scale_color_gradient(low = "blue", high = "red") +
  labs(title = "PCA of Team Performance KPIs",
       x = "Principal Component 1 (PC1)",
       y = "Principal Component 2 (PC2)",
       color = "Wins") +
  theme_minimal()

```


## Checking variance explained by each PC
```{r}
# Calculate the proportion of variance explained by each principal component
variance_explained <- summary(pca_result)$importance[2, ]

# Create a data frame to display it in a table format
variance_df <- data.frame(
  PC = paste0("PC", 1:length(variance_explained)),
  Variance_Explained = variance_explained
)

# Display the table
print(variance_df)


```


```{r}
str(team_runs_wide)
str(team_wickets_wide)
```

## Boxplot of avg_runs_per_over for every venue 

```{r}
# Step 1: Calculate total runs per over for each match
runs_per_over_per_match <- balls %>%
  group_by(match_id, venue, over) %>%
  summarise(runs_per_over = sum(runs_off_bat + extras, na.rm = TRUE)) %>%
  ungroup()

# Step 2: Calculate the number of matches played at each venue
matches_per_venue <- balls %>%
  select(match_id, venue) %>%
  distinct() %>%
  group_by(venue) %>%
  summarise(matches_played = n())

# Step 3: Join the data and calculate average runs per over by dividing by matches played
average_runs_per_over_per_venue <- runs_per_over_per_match %>%
  left_join(matches_per_venue, by = "venue") %>%
  mutate(avg_runs_per_over = runs_per_over / matches_played)

# Step 4: Create a boxplot
ggplot(average_runs_per_over_per_venue, aes(x = venue, y = avg_runs_per_over)) +
  geom_boxplot() +
  labs(title = "Average Runs per Over per Match by Venue",
       x = "Venue",
       y = "Average Runs per Over") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```


## Chord diagram data prep

```{r}
# Chord diagram
# (needed for chordDiagram)
library(circlize)
```

```{r}
# install.packages("circlize")
library(dplyr)

# Calculate total runs per match
runs_per_match <- balls %>%
  group_by(match_id) %>%
  summarise(total_runs = sum(runs_off_bat + extras, na.rm = TRUE), .groups = 'drop')

# Merge total runs with match data
matches_with_runs <- matches %>%
  left_join(runs_per_match, by = c("match_number" = "match_id"))

# Create a new standardized team pair column without altering original team columns
matches_with_runs <- matches_with_runs %>%
  mutate(team_pair = paste(pmin(team1, team2), pmax(team1, team2), sep = " vs "))

# Aggregate results by `team_pair`
matches_between_teams <- matches_with_runs %>%
  group_by(team_pair) %>%
  summarise(
    num_matches = n(),
    total_runs = sum(total_runs, na.rm = TRUE),
    .groups = 'drop'
  )

# View the result
matches_between_teams



```


```{r}
# head(matches_between_teams)
matches_between_teams <- matches %>%
  # Sort team names alphabetically within each row
  mutate(team1 = pmin(team1, team2),
         team2 = pmax(team1, team2)) %>%
  # Group by the unique match-up and count the number of matches
  group_by(team1, team2) %>%
  summarise(num_matches = n(), .groups = 'drop')


```


```{r}
str(matches)
str(balls)

```

## Chord diagram
```{r}
# Chord diagram
# Ensure that matches_between_teams has the columns team1, team2, and num_matches
chordDiagram(matches_between_teams, 
             transparency = 0.5,
             annotationTrack = "grid", 
             preAllocateTracks = list(track.height = 0.1))


# Add titles or labels if needed
title("Matches Played Between Teams in the Tournament")

```


## Saving required tables as RDS files
```{r}

saveRDS(team_performance, "team_performance.rds")
saveRDS(matches_between_teams, "matches_between_teams.rds")
saveRDS(team_runs_wide, "team_runs_wide.rds")
saveRDS(team_wickets_wide, "team_wickets_wide.rds")
saveRDS(pca_scores, "pca_scores.rds")
saveRDS(average_runs_per_over_per_venue,"average_runs_per_over_per_venue.rds")
```

## importing stadium latlong

```{r}
location <- read.csv("data/stadium-latlongs.csv")
str(location)

```

## Adding attributes
```{r}
# Calculate total runs, total wickets, and total matches per venue from balls dataset
venue_stats <- balls %>%
    group_by(venue) %>%
    summarise(
        total_runs = sum(runs_off_bat + extras, na.rm = TRUE),
        total_wickets = sum(!is.na(wicket_type) & wicket_type != ""),
        total_matches = n_distinct(match_id)
    )


venue_stats <- venue_stats %>%
    mutate(venue = sub(",.*", "", venue))


# Merge these statistics with the location table
location <- location %>%
    left_join(venue_stats, by = "venue")



```

```{r}
#head(venue_stats)
str(location)

#summary(location$Latitude)
#summary(location$Longitude)

```




